---
title: "CDS 101 – Final Project Report"
author: "Exponentially Decaying / Nivia, Neel, Justin "
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
---

> Replace this template text with your own writing, keeping the headings and overall structure aligned with the **CDS 101 project rubric**.

# 1. Problem Definition

- Clearly state the main problem or question you are investigating.  
- Explain why it is interesting or important.  
- Define the objective(s) of your analysis.  
- Mention any key assumptions.

-Our project investigates whether weather conditions are related to equine incidents at New York race tracks. The main question we are studying is: How significant is the relationship between equine incidents and weather conditions?

This topic is important because equine deaths and injuries are serious safety concerns in horse racing. By studying incident patterns over time and comparing them with weather conditions, we may be able to identify whether certain weather factors are associated with higher or lower incident counts.

The objective of this analysis is to explore trends in equine incidents from 2009 to 2025, compare incidents across different years, tracks, incident types, seasons, and weather categories, and determine whether weather appears to have a noticeable relationship with incident patterns.

A key assumption is that the incident reports are accurate and consistently recorded. Another assumption is that the weather information connected to each incident is reliable enough to be grouped into categories such as clear, rainy, cloudy, snowy, or unknown.

# 2. Data Acquisition & Description

- Name and source of the dataset(s).  
- How you obtained the data (download link, API, etc.).  
- Brief description of the variables / features (with units if relevant).  
- Approximate size (rows, columns).  
- Any ethical, privacy, or bias considerations (if applicable).

-Equine Death and Breakdown 
-Obtained through downloading from data.gov website.
-There are a lot of variables / features such as Incident Type, Horse, Jockey Driver, Track, and so forth but we will be focusing primarily on two variables, Death or Injury and Weather Conditions.  Death or Injury variable shows if a horse died in an incident or was just injured.  Weather Conditions variable shows the weather conditions at the time of the incident. 
-5,383 rows and 13 columns
-

```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
#Load packages
library(dplyr)
library(lubridate)
library(stringr)

#Load dataset
library(readr)
horses <- read_csv("data/Equine_Death_and_Breakdown.csv")
```

# 3. Data Cleaning & Preprocessing

Describe the steps you took to make the data usable, for example:

- Handling missing values (drop, impute, etc.).  
- Fixing types (dates, numeric vs. categorical).  
- Removing outliers (if done).  
- Creating new features (feature engineering).  
- Any scaling or encoding.

-Fixed the column names, fixed duplicate rows, fixed dates, handled missing values (NA weather conditions)

```{r cleaning, eval=FALSE}
#Fix column names
horses_clean <- horses %>%
  rename(year = Year, incident_date = `Incident Date`, track = Track, inv_location = `Inv Location`, racing_type_description = `Racing Type Description`, division = Division, weather_conditions = `Weather Conditions`, horse = Horse, trainer = Trainer, jockey_driver = `Jockey Driver`, incident_description = `Incident Description`, death_or_injury = `Death or Injury`, incident_type = `Incident Type`)

#Fix dates
horses_clean$incident_date <- mdy(horses_clean$incident_date)

#Get rid of NA entries
horses_clean <- horses_clean %>%
  filter(!is.na(weather_conditions), !is.na(death_or_injury))

#Fix weather_condition column
#USED CHATGPT FOR THIS CODE!!!!!!!!!

horses_clean <- horses_clean %>%
  mutate(
    weather_clean = weather_conditions %>%
      str_to_lower() %>%
      str_trim() %>%
      str_replace_all("[0-9]+", "") %>%        # remove numbers
      str_replace_all("[^a-z ]", "") %>%       # remove symbols
      str_replace_all("couldy|cloudyy|clody", "cloudy") %>%
      str_replace_all("raine", "rain") %>%
      str_squish()
  ) %>%
  
  mutate(
    weather_group = case_when(
      
      # Rain
      str_detect(weather_clean, "rain|shower|drizzle") ~ "Rain",
      
      # Snow
      str_detect(weather_clean, "snow|flurr|sleet") ~ "Snow",
      
      # Storms
      str_detect(weather_clean, "thunder|storm") ~ "Storm",
      
      # Cloudy / Overcast
      str_detect(weather_clean, "cloud|overcast") ~ "Cloudy",
      
      # Clear / Sunny
      str_detect(weather_clean, "clear|sun|fair|dry") ~ "Clear",
      
      # Other weather conditions
      str_detect(weather_clean, "wind|fog|mist|hazy|humid|cold|hot") ~ "Other Weather",
      
      # Missing / junk
      weather_clean %in% c("", "na", "n a", "unc") | is.na(weather_clean) ~ "Unknown",
      
      TRUE ~ "Other"
    )
  )

#Clean death_or_injury column
horses_clean <- horses_clean %>%
  mutate(
    death_or_injury_clean = death_or_injury %>%
      str_to_lower() %>%
      str_trim(),
    
    death_or_injury_clean = case_when(
      str_detect(death_or_injury_clean, "death|dead|euthanasia") ~ "Death",
      str_detect(death_or_injury_clean, "lame|injury|lameness") ~ "Injury",
      TRUE ~ "Other"
    )
  ) 


```

# 4. Exploratory Data Analysis (EDA)

This section maps directly to the **EDA** part of the rubric:

- Summary statistics  
- Visualizations  
- Interpretation connected to the research question  

```{r eda-summary, message=FALSE, warning=FALSE}
# library(dplyr)
# summary(data_clean)
```

```{r eda-plot-example, message=FALSE, warning=FALSE, fig.cap="Example histogram of a numeric variable (EDA rubric)."}
# library(ggplot2)
# ggplot(data_clean, aes(x = some_numeric_variable)) +
#   geom_histogram(binwidth = 5, color = "white") +
#   labs(x = "Some Variable", y = "Count", title = "Distribution of Some Variable")
```

# 5. Visualization Quality and Storytelling

Use this section to satisfy the **Visualization Quality** rubric criterion:

- Explain why your plot types are appropriate.  
- Comment on labels, legends, colors, and overall readability.  
- Mention any steps you took to make plots accessible and interpretable.

# 6. Modeling Approach

Explain how you framed the problem and which models you chose:

- Type of task (regression, classification, etc.).  
- Baseline model or heuristic, if used.  
- Main model(s) chosen and why they are appropriate.

# 7. Model Implementation & Evaluation

This corresponds to the **Model Implementation & Evaluation** rubric criterion.

Describe:

- Features used.  
- Data splitting strategy (train/test or cross-validation).  
- Metrics used (accuracy, F1, RMSE, etc.).  
- Tables/plots summarizing performance.  
- A short interpretation of each metric/plot.

```{r modeling, eval=FALSE}
# Example structure:
# set.seed(123)
# train_index <- sample(seq_len(nrow(data_clean)), size = 0.8 * nrow(data_clean))
# train <- data_clean[train_index, ]
# test  <- data_clean[-train_index, ]
#
# model <- glm(target ~ ., data = train, family = binomial)
# preds <- predict(model, newdata = test, type = "response")
# # compute metrics...
```

# 8. Conclusions & Recommendations

Summarize the key takeaways:

- Answer your original research question(s) directly.  
- Highlight the most important patterns or relationships you found.  
- Discuss limitations (data size, bias, missing variables, etc.).  
- Suggest possible extensions or future work.

# 9. Code Quality & Reproducibility

Briefly document how someone else can reproduce your results:

- Which R scripts or Rmd files to run.  
- Any required R packages.  

```{r session-info, echo=FALSE}
sessionInfo()
```

# 10. References

List any references you used, such as:

- Dataset documentation.  
- Research papers or articles.  
- Tutorials or blog posts.

# Appendix (Optional)

Include extra plots, diagnostic checks, or model comparisons if needed.
